---
output:
  word_document: default
  pdf_document: default
date: "2026-05-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE,
                      message = FALSE
                      )
```


This rmarkdown file contains the data and figures generated for the article titled “No Warning. No Escape. Rethinking Flood Warnings after the Ditwah Cyclone and Devastating Floods in Sri Lanka”, written for the Significance special issue on South and South-East Asia

```{r packages}
library(tidyverse)
library(here)
library(readxl)
library(fable)
library(ceylon)
library(sp)
library(viridis)
library(sf)
library(patchwork)
#library(ggforce)
```
## Figure 2:

The follwoing code produces the map showing the number of people affected by Cyclone Ditwah across Sri Lanka.

The data was obtained from the Ministry of Defence, National Disaster Relief Services Centre (NDRSC), District-wise Current Situation Report (2025-11-21 to 2025-12-31), available at https://ndrsc.gov.lk/web/images/situationreport/2025/current_situation_report_2025_11_21_to2025_12_31.pdf.

```{r}
death <- c(12,12,90,6,9,1,17,1,3,1,254,41,0,64,4,38,0,5,1,82,3,38,2,0,3)
Persons_affected <- c(25760, 76134, 90893, 40101, 328507, 2529, 16049, 
                      2054, 73730, 10617, 195271, 89906, 5028, 95357
                      , 79931, 42692, 2484, 5944, 67347, 61833, 41635,
                     153072, 71647, 91087, 26623 )
data1 <- district|> drop_na()|>
mutate(Death = death,
       Persons_affected = Persons_affected) 

```


```{r fig2, fig.height=5}
p <- ggplot(data1) + 
  geom_sf(aes(fill = Persons_affected),
          color = "white",
          linewidth = 0.2) +
  scale_fill_viridis_c(
    option = "C",
    direction = 1,
    trans = "sqrt",
    name = "People Affected"
  ) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 11, face = "bold"),
    legend.text = element_text(size = 9),
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    plot.caption = element_text(size = 9, color = "grey40"),
     plot.margin = margin(15, 15, 15, 15),
  plot.title.position = "plot"
  ) +
   geom_sf_text(
    data = subset(data1, DISTRICT == "COLOMBO"),
    aes(label = "Colombo"),
    size = 3.5,
    fontface = "bold"
  ) 

print(p)

```

## Figure 3: 

The following code produces the plot of water levels at two gauging stations in the Kelani River basin.

Data Source: Disaster Management Centre (DMC), Sri Lanka, Islandwide Water Level & Rainfall Situation in Major Rivers https://www.dmc.gov.lk/index.php?option=com_dmcreports&view=reports&Itemid=277&report_type_id=6&lang=en

```{r}
data_Kalaniya <- read_csv(here::here("data", "Kalani_river_data.csv"))
colnames(data_Kalaniya)
data_Kalaniya

```
```{r}
data_Kalaniya <- data_Kalaniya %>%
  mutate(Time = Time %>%
           str_replace("\\.", ":") %>%
           mdy_hm())
data_Kalaniya

thresholds <- data_Kalaniya |>
  distinct(`Gauging Station`,
           `Alert Level`,
           `Minor Flood Level`,
           `Major Flood Level`)

event_time <- lubridate::mdy_hm("11/27/2025 12:00 PM")

p <- ggplot(data_Kalaniya, aes(x = Time, y = `Water Level`)) +
  
  # river level line
  geom_line(color = "steelblue", linewidth = 1) +
  
  # threshold lines
  geom_hline(aes(yintercept = `Alert Level`), color = "orange", linetype = "dashed"  )+
  geom_hline(aes(yintercept = `Minor Flood Level`), color = "red", linetype = "dashed") +
  geom_hline(aes(yintercept = `Major Flood Level`), color = "darkred", linetype = "dashed") +
  
  
  # split by station
   facet_wrap(
    ~ factor(`Gauging Station`,
             levels = c("Kithulgala", "Hanwella"), labels = c(
        "Kithulgala" = "Kithulgala - Upstream",
        "Hanwella"   = "Hanwella - Downstream"
      )),
    ncol = 1,
    scales = "free_y"
  ) +
  
  labs(
    x = "Time",
    y = "River water level"
  ) +
  geom_vline(xintercept = as.numeric(event_time),
             linetype = "dashed",
             color = "red",
             linewidth = 1) +
    annotate(
    "label",
    x = event_time,
    y = Inf,
    label = "Actual flood warning \n issued time",
    vjust = 1.5,
    hjust = 1,
    size = 3.5,
    label.size = 0.3,
    fill = "white",
    color = "red"
  ) 
  theme_minimal()
  
print(p)
```

